---
title: "Fitting a cross-lagged panel model"
author: "Sophie W. Berkhout"
date: "5-10-2026"
output: html_notebook
---

### Data set
Data from Van Eersel et al. (2025) made available through DataverseNL. It contains data of employees facing job loss on hours spent job searching and job loss grief at three waves.

The variables in the data set:

-id: id of participant
-x1-3: planned job search hours	(1 = 0 hours; 2 = more than 0 but less than 1 hour; 3 = 1 to 5 hours; 4 = 5 to 10 hours; 5 = 10 to 20 hours; 6 = 20 hours or more)
-y1-3: pre-job loss grief sum score of two times scaled 1-5 (so sum score is between 2-10)

```{r read_data}
dat <- read.csv("jobsearchgrief.csv")
```

### Descriptives

Create boxplots of the scores of both $x$ and $y$ at the three time points.

```{r boxplots}
boxplot(dat$x1, dat$x2, dat$x3)
boxplot(dat$y1, dat$y2, dat$y3)
```

Compute the mean per person across waves for both $x$ and $y$. Do people differ a lot?

```{r}
mean_x <- rowMeans(dat[, c("x1", "x2", "x3")], na.rm = TRUE)
mean_y <- rowMeans(dat[, c("y1", "y2", "y3")], na.rm = TRUE)
hist(mean_x)
hist(mean_y)
```
### Fit the CLPM

Specify the cross-lagged panel model using `lavaan`. To do this, you need to create a character string with `lavaan` code. Then, you can fit the model using the function `sem()`. Note that `lavaan` handles missing data with listwise deletion by default, thus it will throw out any row with missing values. If you want to use all available data, you can specify Full Information Maximum Likelihood with the argument `missing = "ML"`.

You can see the results with `summary()`. Finally, you can get the standardized estimates with the function `standardizedSolution()`.

```{r specify_clpm}
mod_CLPM <- '
  x2 + y2 ~ x1 + y1
  x3 + y3 ~ x2 + y2

  # covariance observed variables first wave
  x1 ~~ y1
  # covariance residuals second and third wave
  x2 ~~ y2
  x3 ~~ y3
'
```

```{r fit_clpm}
if (!require("lavaan")) install.packages("lavaan")
library(lavaan)
fit_CLPM <- sem(mod_CLPM, data = dat, missing = "ML")
summary(fit_CLPM)
standardizedSolution(fit_CLPM)
``` 
- Which regression coefficents are significant?

All except for the cross-lagged regression of x1 to y2 and y1 to x2.

- What standardized cross-lagged effect is the strongest?

The effect of x2 to y3 (-0.177). But there is also a strong effect of y2 to x3 (-0.137).

You can also specify the CLPM with latent mean centering (the alternative formulation).
This way, the intercepts will represent the means of the variables. The regression coefficients should remain the same.

It is a bit of a workaround, so I provide the specification here. It works like this. We create latent variables with a mean of 0 for each variable at each time point. Then we specify the CLPM for these latent variables and say that the observed variables have no residuals themselves. 
Thus, we specify `x2 = intercept + wx2` to create $x_2 = \mu_{x,2} + x^*_2$.

```{r clpm_centered}
mod_CLPM_c <- '
  # within-person centered variables
  wx1 =~ 1*x1
  wx2 =~ 1*x2
  wx3 =~ 1*x3 
  wy1 =~ 1*y1
  wy2 =~ 1*y2
  wy3 =~ 1*y3
  
  # no measurement error in the observed scores
  x1 ~~ 0*x1
  x2 ~~ 0*x2
  x3 ~~ 0*x3
  y1 ~~ 0*y1
  y2 ~~ 0*y2
  y3 ~~ 0*y3
  
  # lagged effects between within-person centered variables
  wx2 + wy2 ~ wx1 + wy1
  wx3 + wy3 ~ wx2 + wy2

  # covariance centered variables first wave
  wx1 ~~ wy1
  # covariances residuals second and third wave
  wx2 ~~ wy2
  wx3 ~~ wx3
'

fit_CLPM_c <- sem(mod_CLPM_c, data = dat, missing = "ML")
summary(fit_CLPM_c)
standardizedSolution(fit_CLPM_c)
``` 

Are the intercepts similar the observed means?

### Extra: `psychonetrics` estimation

Note that this estimation fixes the autoregressive and cross-lagged effects over time.

```{r}
# Specify a design matrix for the lagged effects
Design <- matrix(c(
  "x1", "x2", "x3",
  "y1", "y2", "y3"
), nrow = 2, byrow = TRUE)
rownames(Design) <- c("x", "y")

# Fit panel graphical VAR:
library(psychonetrics)
mod_gvar <- panelgvar(dat, vars = Design, estimator = "FIML",
                      stationary_means = FALSE )
mod_gvar <- runmodel(mod_gvar)
parameters(mod_gvar)
```

### References
Janske van Eersel; Inge Hulshof; Miriam I. Wickham; Geert Smid; Paul Boelen, 2025, "Replication Data for: Reciprocal relations between job loss-related grief reactions and quality of job search behaviour: A cross-lagged analysis", https://doi.org/10.34894/1B1INC, DataverseNL, V1