---
title: "Introduction to the first-order autoregressive model"
author: "Sophie W. Berkhout"
date: "5-10-2026"
output: html_notebook
---

### Generating an AR(1) process
Generating and visualizing model-implied data can often help to understand the
model implied data patterns.The following code simulates data from a first-
order autoregressive model. Run the code so that the function `simulateVAR` is
in your environment. 

```{r sim_ar_function}
simulateAR <- function(nt = 30, c = 3, phi = 0.3, resvar = 1, burnin = 500) {
  # create empty vector of y including burnin time
  # burnin is required to stabilize the process
  y <- numeric(burnin + nt)
  # generate innovations with specified residual variance
  e <- rnorm(burnin + nt, 0, sqrt(resvar))
  # compute y by iterating over time
  # note that since we use y[t - 1] we need to start at t = 2 instead of t = 1,
  # because there is no t = 0.
  for (t in 2:(burnin + nt)) {
    y[t] <- c + phi * y[t - 1] + e[t]
  }
  # create data frame and remove burnin
  df <- data.frame(t = 1:nt, y = y[(burnin + 1):(burnin + nt)] )
  return(df)
}
```

The function takes the following arguments:

- `nt`: The total number of time points, with default $T = 30$.
- `c`: The intercept, with default $c = 3$.
- `phi`: The autoregressive parameter, with default $\phi = 0.3$.
- `resvar`: The residual (i.e., innovation) variance, with default $\sigma^2 = 1$.
- `burnin`: The total number of burnin iterations to use to let the process 
stabilize before saving the data (default = $500$).

The following code runs the function and saves it in a data frame `df`. 

```{r sim_ar_dat}
df <- simulateAR()
```

### Descriptives
Compute descriptive statistics of the generared data.

```{r descriptives}
# your code here
```

### Plot time series
Create a function that plots a time series that takes a data frame as input
with columns `t` and `y`. Also add a vertical line at the observed mean of `y`.

```{r timeseries}
plotTimeSeries <- function(df) {
  # your code here
}
```

### Plot autocorrelation scatter plot
To create a scatter plot for the autocorrelation, you first have to create a
lagged version of `y`. Use the following code chunk to compute the lagged
variable and add this to the data frame `df` with the name `y_lag`.

Inspect the data frame to see whether the lagged variable is the same as the
previous value of `y`.

```{r create_lag}
# your code to add the column y_lag to df
```

Now, you can create a function that plots a scatter plot that takes a data frame
as input with columns `y`, and `y_lag`. Also add a regression line indicating
the autoregressive effect.

```{r scatterplot}
plotScatterAuto <- function(df) {
  # your code here
}
```

```{r plots}
layout(matrix(c(1, 2), nrow = 1), widths = c(3, 2))
plotTimeSeries(df)
plotScatterAuto(df)
layout(1)
```

### Simulating new data sets
You can run the following chunk over and over to simulate new data frames and
visualize them each time. You can adjust the arguments in `simulateAR` to 
see how to process changes.

```{r sim_df_new}
df_new <- simulateAR(nt = 30, c = 3, phi = 0.3, resvar = 1)
layout(matrix(c(1, 2), nrow = 1), widths = c(3, 2))
plotTimeSeries(df_new)
plotScatterAuto(df_new)
layout(1)
```

### Fit an AR(1) model
You have already created a lagged version of `y` to make the autocorrelation
scatter plot. You can now use the same data frame to fit an AR(1) model using
linear regression. Use `y` as the dependent variable and `y_lag` as the
independent variable.

```{r fit_ar}
# your code to fit a linear regression model
mod1 <- lm(y ~ y_lag, df)
mod1
```

Are the parameters (intercept and autoregression) recovered well? You can 
simulate a data set with a larger sample size to confirm. You will have to
create a new lagged column in that case.

Compute the mean from the estimated parameters (recall that
$\mu = \frac{c}{1 - \phi}$).

```{r mean}
# your code here
```

### Alternative AR(1) model
Now run the following chunk. What happens to the parameter values?

```{r fit_ar_alt}
y_mean <- mean(df$y)
cat(paste("The mean of y is: ", round(y_mean, 3)))
df$y_lag_c <- df$y_lag - y_mean
mod2 <- lm(y ~ y_lag_c, df)
mod2
```
