---
title: "Fitting a multilevel VAR(1) model"
author: "Sophie W. Berkhout"
date: "5-10-2026"
output: html_notebook
---

### The data set
Data set from Pavani et al. (2017), made available by openESM (2025). 
14 days, 5 beeps per day

```{r read_data}
dat <- read.csv("0027_pavani.csv")
```

### Creating a lagged variable
We will look at the variable positive affect (`pa`). 
First, we will create a lagged version of this variable.
Make sure that the lagged variable has a missing value for every first beep of
the day. This will also make sure that the last observation of one person is not
linked to the first observation of the next person.

```{r}
# your code here
```

### Descriptives
Compute the mean and autocorrelation of each person's positive affect
and look at their distributions. Do the individuals differ a lot?

Compute the correlations between them. 
How would you interpret these results? It can help to make scatter plots to
visualize the relations between the variables.

```{r descriptives}
# your code here
```

### Time series plot
Select a couple of individuals and plot their time series in one figure.

```{r timeseries}
# your code here
```

Are the time series similar in their means and variances?

### Fit the multilevel AR(1) model

#### Linear mixed effect model with `lme4`
You can estimate a multilevel autoregressive model with a linear mixed effect
model using the `lme4` package. You have to set the lagged variable as the 
predictor and specify a random intercept and random slope. You need another
package to get p-values of the coefficients called `lmerTest`. Just loading
the package should be enough to get the p-values.

Note that we generally prefer to estimate the person means instead of their
intercepts, especially if we are interested the relations between the 
person-specific estimates. Thus, we have to center the lagged variable.

You can extract the person-specific estimates with the function `coef()`.

```{r fit_lme4}
if (!require("lme4")) install.packages("lme4")
library(lme4)
if (!require("lmerTest")) install.packages("lmerTest")
library(lmerTest)
# your code here
```

How do the random effects relate to each other? 

Now compare the random effect estimates to the observed means and 
autocorrelations. Look at their distributions. How do they differ?

Also make scatter plots to visualize the relation between the person-specific
means and autoregressions.
Do these differ from the correlation between the observed mean and
autocorrelation?

```{r lme4_persons}
# your code here
```

Similar to the single-level model, with this kind of univariate estimation
you cannot fit a multivariate multilevel model as it does not allow for the
estimation of the covariance between the innovations.

Therefore, if you are interested in fitting a multilevel vector autoregressive
model, you may want to use a package that can handle multivariate estimation.

#### Multilevel (V)AR estimation with `psychonetrics`
The `psychonetrics` package can fit a multilevel autoregressive model with the
`ml_gvar()` function. For this you do not need a lagged variable, the function
will do this for you. You do, however, need to specify the `dayvar` and 
`beepvar` argument to make sure that the first beep of the day is not regressed
on the last beep of the previous day.

Fit the multilevel AR(1) model using `ml_gvar()`.

```{r}
?ml_gvar
# your code here
```

Unfortunately, I have not found a way to extract the person-specific estimates
using this package. I think it is because it does not estimate the
person-specific effects directly.

The package maintainer (Sacha Epskamp) also has a separate package called
`mlVAR`, which has similar functionality as the `ml_gvar()` function but with a
function named `mlVAR()`. This package has an extra function `resimulate()`, 
which estimates person-specific effects in quite an interesting way.

From the [documentation](https://www.rdocumentation.org/packages/mlVAR/versions/0.7.3/topics/mlVAR) of the `mlVAR` package:
"The `resimulate` method generates simulated data from a fitted `mlVAR` model using person-specific estimated parameters (fixed + random effects). For each person, a VAR time series is simulated using their estimated temporal effects (Beta), contemporaneous precision matrix (Theta), and person-specific means (mu). The output has the same dimensions and missingness pattern as the original input data, making it suitable for posterior predictive checks and model diagnostics."

Why did I still include it in this exercise? The estimation of this function is
really fast. Sometimes researchers are not interested in the person-specific
estimates, and just an estimate of the variance is enough. In that case, this
package is probably the most efficient option.

#### Multilevel (V)AR estimation with `mlts`
If you are interested in the person-specific estimates, there is a package
called `mlts` with this functionality. It uses Bayesian estimation and makes
use of the Stan software. Its estimation takes a little longer, so you have to
be a bit more patient.

You need to use two functions. First, you need to specify what kind of model
you want to estimate using `mlts_model()`. Then you can fit the model using
`mlts_fit()`.

Note that for the specification in `mlts_fit()` the `beep` requires a different
formulation than the `beepvar` argument of the `psychonetrics` package. Here,
the beep number has to go up across days. For example, if there are five beeps
per day, then the beep indicator of the first beep of the second day would be
6 instead of 1. Luckily, the data already came with such a variable, called
`counter`.

You also need to specify how you want to deal with the night gap with the
argument `n_overnight_NAs`. You have to indicate the number of NA's that you
want it to insert between the last beep of a day and the first beep of the next
day. A higher number means that the effect of the ast beep of a day and the 
first beep of the next day is smaller. The model estimates missing data using 
Bayesian estimation, so you do not want to set this too high, which would make
it harder to estimate the missing values.

Finally, to extract the person-specific estimates you need to set the argument
`monitor_person_pars` to `TRUE`.

A tutorial for a multilevel AR(1) model is provided here: 
https://cran.r-project.org/web/packages/mlts/vignettes/manifest_ar.html

```{r fit_multilevel}
if (!require("mlts")) install.packages("mlts")
library(mlts)
?mlts_model
?mlts_fit
# your code here
```

A pro of this package is that it offers multivariate estimation as well as
estimation of predictor effects on the between-level. For example, you could
estimate whether neuroticism (measured once) can predict the person-specific autoregressive effect estimates.

Compare the estimates of each of these three estimation techniques. Are they
similar? Are there any differences?

### Extra exercise
Fit a multilevel vector autoregressive model with either the `psychonetrics` or
the `mlts` package. You can choose any of the variables from the data set.


### References
openESM. (2025). 0027_pavani (Version 1.0.0) [Data set]. Zenodo. https://doi.org/10.5281/zenodo.17347893

Pavani, J., Le Vigouroux, S., Kop, J., Congard, A., & Dauvier, B. (2017). A network approach to affect regulation dynamics and personality trait–induced variations: Extraversion and neuroticism moderate reciprocal influences between affect and affect regulation strategies. European Journal of Personality, 31(4), 329–346. https://doi.org/10.1002/per.2109