---
title: "Fitting a multilevel VAR(1) model"
author: "Sophie W. Berkhout"
date: "5-10-2026"
output: html_notebook
---

### The data set
The data set we will use is from Pavani et al. (2017), made available by openESM (2025).
It has the following variables:

- `id`: participant ID
- `day`: day number per individual
- `beep`: beep number per day
- `counter`: measurement occasion per individual
- `pa`: positive affect (average score between 1-5)
- `na`: negative affect (average score between 1-5)
- `appreciation`: I have taken time to savor a moment of the day (scale 1-5)

```{r read_data}
dat <- read.csv("0027_pavani.csv")
```

### Creating a lagged variable
We will look at the variable positive affect (`pa`). 
First, we will create a lagged version of this variable.
Make sure that the lagged variable has a missing value for every first beep of
the day. This will also make sure that the last observation of one person is not
linked to the first observation of the next person.

Hint: The code from the $N=1$ data should work here as well.

```{r}
dat$pa_lag <- NA
for (i in 2:nrow(dat)) {
  dat$pa_lag[i] <- ifelse((dat$beep[i] - dat$beep[i - 1]) == 1, dat$pa[i - 1], NA)
}
```

### Descriptives
Compute the mean and autocorrelation of each person's positive affect
and look at their distributions. Do the individuals differ a lot?

Also compute the correlations between them. 
How would you interpret these results? 

```{r descriptives}
df_descriptives <- dat %>%
  group_by(id) %>%
  summarise(mean = mean(pa, na.rm = TRUE),
            autocorrelation = cor(pa, pa_lag, use = "complete.obs")) %>%
  ungroup()

hist(df_descriptives$mean)
hist(df_descriptives$autocorrelation)

cor(df_descriptives$mean, df_descriptives$autocorrelation)
```
There is a small negative correlation between the average positive affect and the autocorrelation, which means that people who have a higher than average positive affect tend to have a smaller autocorrelation.

### Time series plot
Select a couple of individuals and plot their time series in one figure.

```{r timeseries}
plot(dat[dat$id == unique(dat$id)[1], "counter"], dat[dat$id == unique(dat$id)[1], "pa"], type = "l", ylim = c(1, 5))
lines(dat[dat$id == unique(dat$id)[2], "counter"], dat[dat$id == unique(dat$id)[2], "pa"], col = "blue")
lines(dat[dat$id == unique(dat$id)[3], "counter"], dat[dat$id == unique(dat$id)[3], "pa"], col = "red")
lines(dat[dat$id == unique(dat$id)[4], "counter"], dat[dat$id == unique(dat$id)[4], "pa"], col = "green")
lines(dat[dat$id == unique(dat$id)[5], "counter"], dat[dat$id == unique(dat$id)[5], "pa"], col = "orange")
```

Are the time series similar in their means and variances?
Some people are similar to each other, but there are also some who are quite different (e.g., the orange en green lines).

### Fit the multilevel AR(1) model

#### Linear mixed effect model with `lme4`
You can estimate a multilevel autoregressive model with a linear mixed effect
model using the `lme4` package. You have to set the lagged variable as the 
predictor and specify a random intercept and random slope. You need another
package to get p-values of the coefficients called `lmerTest`. Just loading
the package should be enough to get the p-values.

Note that we generally prefer to estimate the person means instead of their
intercepts, especially if we are interested the relations between the 
person-specific estimates. Thus, the preference is to center the lagged variable.
However, in this case

You can extract the person-specific estimates with the function `coef()`.

```{r fit_lme4}
if (!require("lme4")) install.packages("lme4")
library(lme4)
if (!require("lmerTest")) install.packages("lmerTest")
library(lmerTest)

# grand mean centering
dat$pa_lag_c <- dat$pa_lag - mean(dat$pa_lag, na.rm = TRUE)

# person mean centering
dat$pa_lag_c <- with(
  dat, pa_lag - ave(pa, id, FUN = function(x) mean(x, na.rm = TRUE))
)

mod_lm <- lmer(pa ~ pa_lag_c + (1 + pa_lag_c | id), data = dat)
summary(mod_lm)
```

How do the random effects relate to each other? 

You can find the correlation under 'Random effects:', in the row of `pa_lag_c.` The correlation between the means (intercepts) and the autoregressions is $-0.08$. So there is barely any relation.

Plot the estimated means against the estimated autoregressions.

```{r lme4_persons}
# your code here
re <- coef(mod_lm)$id
plot(re$`(Intercept)`, re$pa_lag_c)
```

Similar to the single-level model, with this kind of univariate estimation
you cannot fit a multivariate multilevel model as it does not allow for the
estimation of the covariance between the innovations.

Therefore, if you are interested in fitting a multilevel vector autoregressive
model, you may want to use a package that can handle multivariate estimation.

#### Multilevel (V)AR estimation with `psychonetrics`
The `psychonetrics` package can fit a multilevel autoregressive model with the
`ml_gvar()` function. For this you do not need a lagged variable, the function
will do this for you. You do, however, need to specify the `dayvar` and 
`beepvar` argument to make sure that the first beep of the day is not regressed
on the last beep of the previous day.

Fit the multilevel AR(1) model using `ml_gvar()`.

```{r}
?ml_gvar
# your code here
mod_gar <- ml_gvar(dat, vars = "pa", idvar = "id",
                   dayvar = "day", beepvar = "beep")
mod_gar <- runmodel(mod_gvar)
parameters(mod_gar)
```

Unfortunately, I have not found a way to extract the person-specific estimates
using this package. I think it is because it does not estimate the
person-specific effects directly.

The package maintainer (Sacha Epskamp) also has a separate package called
`mlVAR`, which has similar functionality as the `ml_gvar()` function but with a
function named `mlVAR()`. This package has an extra function `resimulate()`, 
which estimates person-specific effects in quite an interesting way.

From the [documentation](https://www.rdocumentation.org/packages/mlVAR/versions/0.7.3/topics/mlVAR) of the `mlVAR` package:
"The `resimulate` method generates simulated data from a fitted `mlVAR` model using person-specific estimated parameters (fixed + random effects). For each person, a VAR time series is simulated using their estimated temporal effects (Beta), contemporaneous precision matrix (Theta), and person-specific means (mu). The output has the same dimensions and missingness pattern as the original input data, making it suitable for posterior predictive checks and model diagnostics."

Why did I still include it in this exercise? The estimation of this function is
really fast. Sometimes researchers are not interested in the person-specific
estimates, and just an estimate of the variance is enough. In that case, this
package is probably the most efficient option.

#### Multilevel (V)AR estimation with `mlts`
If you are interested in the person-specific estimates, there is a package
called `mlts` with this functionality. It uses Bayesian estimation and makes
use of the Stan software. Its estimation takes a little longer, so you have to
be a bit more patient.

You need to use two functions. First, you need to specify what kind of model
you want to estimate using `mlts_model()`. Then you can fit the model using
`mlts_fit()`.

Note that for the specification in `mlts_fit()` the `beep` requires a different
formulation than the `beepvar` argument of the `psychonetrics` package. Here,
the beep number has to go up across days. For example, if there are five beeps
per day, then the beep indicator of the first beep of the second day would be
6 instead of 1. Luckily, the data already came with such a variable, called
`counter`.

You also need to specify how you want to deal with the night gap with the
argument `n_overnight_NAs`. You have to indicate the number of NA's that you
want it to insert between the last beep of a day and the first beep of the next
day. A higher number means that the effect of the ast beep of a day and the 
first beep of the next day is smaller. The model estimates missing data using 
Bayesian estimation, so you do not want to set this too high, which would make
it harder to estimate the missing values.

Finally, to extract the person-specific estimates you need to set the argument
`monitor_person_pars` to `TRUE`.

A tutorial for a multilevel AR(1) model is provided here: 
https://cran.r-project.org/web/packages/mlts/vignettes/manifest_ar.html

```{r fit_multilevel}
if (!require("mlts")) install.packages("mlts")
library(mlts)
?mlts_model
?mlts_fit
# your code here
ar1_model <- mlts_model(q = 1)
mod_mlts <- mlts_fit(
  model = ar1_model,
  data = dat,
  id = "id",
  ts = "pa",
  iter = 2000,
  monitor_person_pars = TRUE,
  n_overnight_NAs = 6
)
```

A pro of this package is that it offers multivariate estimation as well as
estimation of predictor effects on the between-level. For example, you could
estimate whether neuroticism (measured once) can predict the person-specific autoregressive effect estimates.

Compare the estimates of each of these three estimation techniques. Are they
similar? Are there any differences?

### Extra exercise
Fit a multilevel vector autoregressive model with either the `psychonetrics` or
the `mlts` package. You can choose any of the variables from the data set.

```{r extra}
# mlts estimation
var1_model <- mlts_model(q = 2)
mod_mlts <- mlts_fit(
  model = ar1_model,
  data = dat,
  id = "id",
  ts = c("pa", "na"),
  iter = 2000,
  monitor_person_pars = TRUE,
  n_overnight_NAs = 6
)

# psychonetrics estimation
mod_gvar <- ml_gvar(dat, vars = c("pa", "na"), idvar = "id",
                    dayvar = "day", beepvar = "beep")
mod_gvar <- runmodel(mod_gvar)
parameters(mod_gvar)
```

### References
openESM. (2025). 0027_pavani (Version 1.0.0) [Data set]. Zenodo. https://doi.org/10.5281/zenodo.17347893

Pavani, J., Le Vigouroux, S., Kop, J., Congard, A., & Dauvier, B. (2017). A network approach to affect regulation dynamics and personality trait–induced variations: Extraversion and neuroticism moderate reciprocal influences between affect and affect regulation strategies. European Journal of Personality, 31(4), 329–346. https://doi.org/10.1002/per.2109