---
title: "Fitting the VAR(1) model to empirical data"
author: "Sophie W. Berkhout"
date: "5-10-2026"
output: html_notebook
---

### Data from Peter Groot
For this exercise you will use the same data set as the previous exercise.
The file 'data_groot_prepped.csv' contains the data with lagged variables, so
you do not have to compute these again.

```{r read_data}
dat <- read.csv("data_groot_prepped.csv")
```

We are going to look at the cross-lagged relations between `selfdoubt` and `tired`.

### Plots
First make a time series plot with both variables in the same plot. Give the
lines a different colour. Can you see similar patterns?

```{r plots_timeseries}
# your code here
```

Similar to an autocorrelation scatter plot, you can plot the cross-lagged relations; 
one with $y_t$ versus $x_{t-1}$ and one with $x_t$ versus $y_{t-1}$.

Make cross-lagged relation scatter plots for `self-doubt` versus `tired`, and vice versa.

```{r plots_cross_lagged}
# your code here
```

### Fit the VAR(1) model
First, estimate two univariate models, one with `selfdoubt` as outcome and one 
with `tired` as outcome, and the lagged versions of both variables as predictors.
Center the predictors if you want the intercept to represent the mean of the process.

```{r univariate}
# univariate model outcome selfdoubt

# univariate model outcome tired

```

Answer the following questions:
- Are the autoregressive effects similar as the AR(1) model?
- Is there a cross-lagged effect from `selfdoubt` to `tired`?
- Is there a cross-lagged effect from `tired` to `selfdoubt`?
- How would you interpret the cross-lagged effects?

Unfortunately, with univariate estimation we cannot always estimate the 
covariance between the residuals. As is the case here: because of different
missing value patterns, the residuals of the two models do not match.
A solution would be to remove all incomplete cases in the full data set (e.g., using `na.omit`). You can then extract the residuals, for example using the `residuals()` function from the `stats` package, and compute the covariance/correlation between them.

Remove all missing cases from the data and fit the two univariate AR(1) models again. Compute the covariance between the residuals of the two models.
```{r cov_residuals}
# your code here
```

A more elegant approach would be to estimate the VAR(1) model multivariately.
The `vars` package is unfortunately not an option here, as it cannot handle missing data.
We can, however, use the `var1()` function from the `psychonetrics` package for this.
Conveniently, it has build-in functionality to handle lags and separate days and beeps
in ESM data, through the arguments `dayvar` and `beepvar`, for which you have to
provide to column names that indicate the days and the beep numbers, respectively.
After you specified the model with the `var1()` function, you can fit the model with the `runmodel()` function. Finally, you can extract the parameter estimates with the `parameters()` function.

The package has very extensive documentation, see e.g., https://psychonetrics.org/var1.html.

Fit a multivariate VAR(1) model.

```{r vars}
if (!require("psychonetrics")) install.packages("psychonetrics")
library(psychonetrics)
?var1
```

When looking at the output of the `parameters()` function, you will see the 
estimated means (`mu`), the regression coefficients (`beta`), and the
covariance matrix of the residuals (`sigma_zeta`).

Answer the following questions:
- Are the means and regression coefficients similar to the univariate estimates?
- Is the estimated residual covariance similar to the one you computed from the univariate residuals?

